Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-23 paper.tex
Updated refs for JLFs section

Enumerating molecules is a combinatorial problem that has fascinated chemists,
computer scientists and mathematicians alike for more than a century. It is one
important risk-reduction element in still being able to encode closely-related
molecules and for keeping the space complexity of databases under (algorithmic)
control. Indeed, while trying to solve the problem of counting the isomers of
paraffin structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley, Polya and
others. Even terms that are widely-used nowadays like graph and tree were
originally coined in a chemistry context (Chapters 1 and 8 in [1]). More than
four decades ago, Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others
from Stanford University started developing codes to explicitly enumerate
molecules. Their project, named DENDRAL, was funded by NASA with the goal of
developing an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer science were
developed. DENDRAL is generally quoted in artificial intelligence textbooks as a
pioneer project that produced the first expert system.

@@

The major difficulty with enumeration is that the \emph{in silico} representation of
molecules, the so-called molecular graphs, are labeled objects (i.e. labeled
graphs), while the atoms of a molecule are of course not uniquely labeled. The
mathematical concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. When enumerating
molecular structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for molecular
graphs due to the fact that molecules belong to a restricted class of graphs
known as \emph{bounded valence graphs}; however, that same restriction prevents
applying to a molecular graph the polynomial delay enumeration results obtained
for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that molecules
can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate natural
compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries like GDB-13
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers [3] or tautomers [4]; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
(UCSF) are valuable resources to search for new drugs. For instance
the well known ligand-based design approach uses these
databases along with structure similarity tools to discover new
compounds binding to biological targets (protein or nucleic
acid). Compounds retrieved through ligand-based design, however, must
already be present in databases, which is a limitation of the
technique. To overcome that limitation, algorithms have recently been
developed to enumerate all compounds in the chemical space matching
circular [5] or path fingerprints [6], these fingerprints being
popular tools to measure similarity. Enumerating molecules from
fingerprints, or more generally enumerating molecules from molecular
descriptors, is still very much under development. At stake is the
design of new chemicals or drugs that have not yet been synthesized.

Another area where enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in [1] several network enumeration techniques have been
developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
chemoinformatics and bioinformatics.

>>>>

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike for more than a
century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays like graph
and tree were originally coined in a chemistry context (Chapters 1 and
8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
Carl Djerassi, Edward Feigenbaum and others from Stanford University
started developing codes to explicitly enumerate molecules. Their
project, named DENDRAL, was funded by NASA with the goal of developing
an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer
science were developed. DENDRAL is generally quoted in artificial
intelligence textbooks as a pioneer project that produced the first
expert system.

@@

The major difficulty with enumeration is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. When enumerating molecular
structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in
\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently \cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries like GDB-13
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
(UCSF) are valuable resources to search for new drugs. For instance
the well known ligand-based design approach uses these databases along
with structure similarity tools to discover new compounds binding to
biological targets (protein or nucleic acid). Compounds retrieved
through ligand-based design, however, must already be present in
databases, which is a limitation of the technique. To overcome that
limitation, algorithms have recently been developed to enumerate all
compounds in the chemical space matching circular \cite{faulon2003} or
path fingerprints \cite{fujiwara2008}, these fingerprints being
popular tools to measure similarity. Enumerating molecules from
fingerprints, or more generally enumerating molecules from molecular
descriptors, is still very much under development. At stake is the
design of new chemicals or drugs that have not yet been synthesized.

Another area where enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
chemoinformatics and bioinformatics.
